---
title: "CirComPara: detection and analysis of circular RNAs"
date: "`r Sys.Date()`"  
output: 
  html_document: 
    keep_md: no
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: 
      collapsed: false
      smooth_scroll: true
    theme: "readable"
---

# CirComPara results summary
CirComPara analysis results are presented in this page. Different sections display aggregated data in tables and figures. Raw and integrated results were saved in comma separated (CSV) files. 

```{r global_options, include=FALSE}
## variables to be defined in parent frame
# results.dir
# circrnas.gtf.file
# combined_gtf_file# 
# gene_meta_file
# gene_fpkm_file
# min_methods ## minimum number of methods detecting a circRNA to define it "reliable"
# min_reads

dir.create(path = results.dir, showWarnings = F, recursive = T)
knitr::opts_chunk$set(fig.width = 6, 
                      fig.height = 4, 
                      dev = c('svg', 'png'), 
                      fig.pos = "center",
                      fig.path = file.path(results.dir, paste0('Figs', .Platform$file.sep)),
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
library(knitr)
library(data.table)
library(ggplot2)

```

```{r utility_functions, echo=FALSE, include=FALSE}
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

# Compute label for a circRNA according to gene overlaps
#
# This function serves to categorize backsplices according to the positions
# relative to gene/exons.
# An "error region" of the exon/backsplice coordinates could be considered
# by setting flank > 0
# We define three backsplice categories:
# 1. Intergenic: both start and end are outside any gene
# 2. Intronic: both end are in intron(s)
# 3. Exonic: either start or end is in exon(s)
#
# - backsplice: a two elements integer array with start and end of the circRNA
# - exon:       a two elements integer array with start and end of the exon
# - gene:       a two elements integer array with start and end of the gene
# - flank:      the number of flanking bases allowed for the backspice coordinates
#
label.backsplice <- function(backsplice, exon, gene, flank = 0){
  
  gene_start <- gene[1] - flank
  gene_end <- gene[2] + flank
  exon_start <- exon[1] - flank
  exon_end <- exon[2] + flank
  circ_start <- backsplice[1]
  circ_end <- backsplice[2]
  
  category <- "undefined"
  if(exon_start < 0){
    category <- "intergenic"
  }else{
    if(circ_start >= gene_start | circ_end <= gene_end){ # are we inside a gene?
      category <- "intronic"
      if(circ_start >= exon_start | circ_end <= exon_end){ # are we within an exon?
        category <- "exonic"
      }
    }else{
      # we are not inside a gene, but an exon is overlapped
      if(circ_start < gene_start & circ_end > gene_end){ # just double check we are completely outside a gene
        category <- "intergenic_spanning_gene"
      }
    }
  }
  category
}

# Refine labelling for a circRNA
#
# - labels_list: the list of labels assigned to i circRNA ID
#
refine_label <- function(labels_list){
  
  if("exonic" %in% labels_list){
    label <- "exonic"
  }else{
    if("intronic" %in% labels_list){
      label <- "intronic"
    }else{
      if("intergenic" %in% labels_list){
        label <- "intergenic"
      }else{
        label <- paste0(sort(unique(labels_list)), collapse = ",")
      }
    }
  }
  label
}
```

```{r sample_summary, echo=FALSE, include=FALSE}
"TODO"
## meta <- fread(meta.file)
## read.preprocessing
```

<!-- 
TODO: Considered samples: list (+ altri dati?)
(insert summary per sample? E.g. total reads or other statistics that could describe sample differences?)
-->

```{r load_circrnas, echo=FALSE, include=FALSE}
## read circRNA results: filter low expressed (less than min_reads reads) circRNAs
colClasses <- c("factor", "factor", "character", "integer", "integer", "integer", "factor", "character", "character")
circrnas.gtf <- fread(circrnas.gtf.file, data.table = T, colClasses = colClasses)[V6 >= min_reads]
V9pattern <- '.*gene_id "([^"]*)".*transcript_id "([^"]*)".*'
V9newCols <- c("gene_id", "transcript_id")
exists_circrnas <- FALSE
if(nrow(circrnas.gtf) > 0){
  exists_circrnas <- TRUE
  circrnas.gtf[, `:=` (sample_id = sub('.*sample_id "([^"]*)".*', "\\1", V9))
               ][, (V9newCols) := tstrsplit(sub(V9pattern, "\\1@\\2", V9), "@", fixed=TRUE)
                 ][, V9 := NULL]
  
  # if(fix_testrealign){
  #   setkey(circrnas.gtf, V2)
  #   circrnas.gtf["testrealign", V4 := V4-1]
  #   circrnas.gtf["testrealign", `:=`(gene_id = paste0(V1, ":", V4, "-", V5, ":", V7), 
  #                                 transcript_id = paste0(V1, ":", V4, "-", V5, ":", V7, ".", sample_id))]
  # }
  
  # ## read circRNA-method per sample normalization factors
  # norm.factors.file <- "/home/enrico/Scrivania/QKI_anhal/analysis/circRNA_collection/circrna_maps_counts.txt"
  # norm.factors <- as.data.table(scan(file = norm.factors.file, what = list("character", "integer"), sep = "\n"))
  # norm.factors[, `:=` (sample_id = sub('.*samples/([^/]*).*', "\\1", V1),
  #                      method = tolower(gsub("_out|_", "", sub('.*circRNAs/([^/]*).*', "\\1", V1))),
  #                      norm_factor = as.integer(V2))][, `:=` (V1 = NULL, V2 = NULL)]
  # setkey(x = norm.factors, sample_id, method)
  # ## apply normalization factors: define normalized expression as RPM (Reads Per Million mapped reads)
  # setkey(x = circrnas.gtf, sample_id, V2)
  # circrnas_expression <- norm.factors[circrnas.gtf][, RPM := (V6*10^6)/norm_factor]
  
  ## normalize circrna reads by the amount of backsplice Reads in method (Per Million reads - RPM)
  circrnas.gtf[, RPM := (V6/sum(V6))*10^6, by = .(sample_id, V2)]
  ## compute median of normalized value
  circrna.rpms <- data.table::dcast(data = circrnas.gtf, formula = sample_id + gene_id + V1 + V4 + V5 + V7 ~ V2, value.var = "RPM", fill = 0)
  circrna.rpms$medRPM <- apply(X = circrna.rpms[, 7:ncol(circrna.rpms), with = F], MARGIN = 1, FUN = median)
}else{
  circrnas.gtf[, `:=`(RPM="", gene_id="")]
  circrna.rpms <- circrnas.gtf
  circrna.rpms.genes <- circrna.rpms
  circrna.rpms.genes[, `:=`(circ_id="", gene_ids="", gene_names="", label="", sample_id="", chr="", V1="", start="", V4="", 
                            end="", V5="", strand="", V7="", medRPM="")]
}

```

```{r load_circrna_gene_overlaps, echo=FALSE, include=FALSE}
## retrieve and attach gene symbols and gene_ids: use only exons hit by the backsplice coordinates 
colClasses <- c("factor", "factor", "character", "integer", "integer", "integer", "factor", "character", "character",
                "factor", "factor", "character", "integer", "integer", "character", "factor", "character", "character")
#colToDrop <- c("V3", "V8", "V11", "V12", "V15", "V17")
#combined_gtf <- fread(paste0('zcat ', combined_gtf_file), data.table = T, colClasses = colClasses, drop = colToDrop)
combined_gtf <- fread(paste0('zcat ', combined_gtf_file), data.table = T, colClasses = colClasses)
exists_gene_overlaps <- FALSE
if(nrow(combined_gtf) > 0){
  exists_gene_overlaps <- TRUE
  # if(fix_testrealign){
  #   setkey(combined_gtf, V2)
  #   combined_gtf["testrealign", V4 := V4-1]
  # }
  combined_gtf[, `:=`(gene_id = sub('.*gene_id "([^"]*)".*', "\\1", V18), 
                      gene_name = ifelse(grepl('gene_name', V18, fixed = T), 
                                         sub('.*gene_name "([^"]*)".*', "\\1", V18), 
                                         "."),
                      chr = V1, circ_start = as.integer(V4), circ_end = as.integer(V5), strand = V7,
                      exon_start = as.integer(V13), exon_end = as.integer(V14))
               ][, `:=`(V1 = NULL, V4 = NULL, V5 = NULL, V7 = NULL, V13 = NULL, V14 = NULL)]
  combined_gtf[, `:=`(gene_start = min(exon_start), gene_end = max(exon_end)), by = gene_id]
  
  circ_to_gene <- combined_gtf[, .(counts = .N), by = .(chr, circ_start, circ_end, strand, 
                                                        exon_start, exon_end, gene_id, gene_name, 
                                                        gene_start, gene_end)][, counts := NULL]
  
  
  
  circ_to_gene[, rowId := .I]
  circ_to_gene[, label := label.backsplice(c(circ_start, circ_end), 
                                           c(exon_start, exon_end), 
                                           c(gene_start, gene_end), 
                                           flank = 1),
               by = rowId]
  circ_to_gene[, `:=` (rowId = NULL)]
  
  # circ_to_gene[, .(circ_id = paste0(chr, ":", circ_start, "-", circ_end, ":", strand),
  #                  gene_id, gene_name, label)][
  #                    , .(label = paste0(sort(unique(label)), collapse = "|")), by = .(circ_id, gene_id, gene_name)][, .N, by = label]
  
  circ_to_genes <- circ_to_gene[, .(circ_id = paste0(chr, ":", circ_start, "-", circ_end, ":", strand), 
                                    gene_id, gene_name, label)][
                                      , .(label = refine_label(label)), by = .(circ_id, gene_id, gene_name)][
                                        , .(gene_ids = paste0(unique(gene_id), collapse = "|"),
                                            gene_names = paste0(unique(gene_name), collapse = "|"),
                                            label = paste0(unique(label), collapse = "|")), 
                                        by = circ_id]
  
}

if(exists_circrnas & exists_gene_overlaps){
  setkey(circrna.rpms, gene_id)
  setkey(circ_to_genes, circ_id)
  circrna.rpms.genes <- circ_to_genes[circrna.rpms]
  
  ## split results per sample
  ## NB: use split function from data.table >= 1.9.7
  circrnas_expression.samples <- split(circrna.rpms.genes, by = "sample_id", keep.by = F, drop = T)
  
  ## save tables: one per sample
  save.sample.table <- function(sample, outdir){
    table.file <- file.path(outdir, paste0(sample, "_circrna_norm_expression.csv"))
    write.csv(x = circrnas_expression.samples[[sample]], file = table.file, row.names = F)
    table.file
  }
  
  sample.table.file.names <- sapply(X = names(circrnas_expression.samples), 
                                    FUN = save.sample.table, 
                                    outdir = results.dir)
}else{
  sample.table.file.names <- "No circRNAs detected"
}
```

# Detected circRNA results

Backsplices were detected using `r circrnas.gtf[, .N, by = V2][, .N]` programs in parallel: `r paste0(sort(circrnas.gtf[, .N, by = V2][, V2]), collapse = ", ")`

The detected circRNAs were saved in the following files:
```{r sample_circrna_table_files, echo=FALSE, include=TRUE}
kable(data.frame(sample_id = names(sample.table.file.names), file = sample.table.file.names, row.names = NULL))#, 
#      caption = "Files reporting the detected circRNAs in each sample")
```

```{r reliable_circrnas, echo=FALSE, include=TRUE}
############# CircRNA expression per sample ###########
## get the list of circRNAs detected by min_methods or more methods
circrnas.morethan2methods <- circrnas.gtf[, .N, 
                                          by = .(gene_id, V2)][
                                            , N := NULL][
                                              , .(n_methods = .N), 
                                              by = gene_id][
                                                n_methods >= min_methods, .(gene_id)]

## subset the circRNA expression table by the circRNAs detected by at least min_methods methods and
## with normalized median expression > 0
setkey(circrna.rpms.genes, circ_id)
setkey(circrnas.morethan2methods, gene_id)
circrnas.morethan2methods.medrpm <- circrna.rpms.genes[circrnas.morethan2methods][
  , .(circ_id, gene_ids, gene_names, label, sample_id, chr = V1, start = V4, end = V5, strand = V7, medRPM)][medRPM > 0]

circrnas.morethan2methods.medrpm.table <- circrnas.morethan2methods
if(nrow(circrnas.morethan2methods) > 0 ){
  ## reshape with one expression column per sample 
  circrnas.morethan2methods.medrpm.table <- data.table::dcast(data = circrnas.morethan2methods.medrpm, 
                                                              formula = circ_id + gene_ids + gene_names + label + chr + start + end + strand ~ sample_id,
                                                              value.var = "medRPM", fill = 0)
  
  # ## add a useful mean expression column
  # circrnas.morethan2methods.medrpm.table$avg_xpr <- apply(X = circrnas.morethan2methods.medrpm.table[
  #   , 9:ncol(circrnas.morethan2methods.medrpm.table)], 
  #   MARGIN = 1, FUN = mean)
}
## save table
write.csv(x = circrnas.morethan2methods.medrpm.table, 
          file = file.path(results.dir, "circRNA_expression_per_sample.csv"), 
          row.names = F)

## TODO: save in HTML table with links
```

## CircRNAs detected per method

```{r show_circrnas_per_method, echo=FALSE, include=TRUE}

circ.per.method <- circrnas.gtf[V6 >= min_reads][, .N, by = .(gene_id, V2)]
#circ.per.method[, .N, by = V2][, Method := V2][order(-N), .(Method, N)]
non.redundant <- circ.per.method[, .N, by = gene_id][, .N]
## number of circRNAs per method
kable(rbind(circ.per.method[, .N, by = V2][, Method := V2][order(-N), .(Method, "Tot circRNAs" = N)], 
            list("Non-redundant", non.redundant)))

############# Barplot number of circRNAs per methods (sum over all samples) ###########
default.font.size <- 12
## select only circRNAs with >= 2 reads (not normalized values)
circByMethod.barplot <- ggplot(data = circ.per.method, mapping = aes(x = V2, fill = V2)) + 
  geom_bar(width = .7, colour="black") + 
  geom_text(stat='count', aes(label=..count..), vjust=-0.3, size=default.font.size*(5/14)) + 
  theme_bw() + xlab("CircRNA detection method") + ylab("CircRNA number") +
  ggtitle(paste0("CircRNAs detected with >=", min_reads, "reads by each method"))

## compute limits to remove space from x-axis. Add a 10% increase to the top limit
circByMethod.barplot.limits <- c(0, max(circ.per.method[, .N, by = V2][, N])*1.1)
circByMethod.barplot + guides(fill=FALSE) + 
  theme(text = element_text(size=default.font.size)) + 
  scale_y_continuous(expand = c(0, 0), limits = circByMethod.barplot.limits) + 
  scale_fill_manual(values = rainbow(circ.per.method[, .N, by = V2][, .N]))

```

### CircRNAs shared by methods

```{r show_circrnas_shared, echo=FALSE, include=TRUE}

############# Matrix of methods' shared circRNA counts ###########

shared.counts.table <- crossprod(x = table(circ.per.method[
  , `:=`(N = NULL, method = V2, V2 = NULL)]))
## save table
write.csv(x = data.frame(shared.counts.table), file = file.path(results.dir, "methods_shared_circRNA_counts.csv"), row.names = F)

kable(data.frame(shared.counts.table), caption = "")

############# Barplot number of circRNAs shared by methods ###########

sharedByMethod.barplot <- ggplot(data = circ.per.method[, .(shared_by = .N), by = gene_id], 
                                 mapping = aes(x = factor(shared_by), fill = factor(shared_by))) + 
  geom_bar(width = .7, color = "black") + 
  geom_text(stat='count', aes(label=..count..), vjust=-0.3, size=default.font.size*(5/14)) + 
  theme_bw() + xlab("Number of methods") + ylab("CircRNA number") +
  ggtitle(paste0("Number of circRNAs (with >= ", min_reads, " reads)\ncommonly detected by the methods"))
## compute limits to remove space from x-axis. Add a 10% increase to the top limit
sharedByMethod.barplot.limits <- c(0, max(circ.per.method[, .(shared_by = .N), by = gene_id][, .N, by=shared_by][, N])*1.1)
sharedByMethod.barplot + guides(fill=FALSE) + 
  theme(text = element_text(size=default.font.size)) + 
  scale_y_continuous(expand = c(0, 0), limits = sharedByMethod.barplot.limits) + 
  scale_fill_brewer(direction = 1)

```

## CircRNAs detected in total

For each method, at least `r min_reads` reads were required to consider a circRNAs as detected. Further, circRNAs detected by at least `r min_methods` methods are considered more reliable.

With at least `r min_reads` reads | With at least `r min_reads` reads and `r min_methods` methods
---------------: | -----------------------------------------:
`r circrnas.gtf[RPM > 0, .N, by = .(gene_id)][, .N]` | `r circrnas.morethan2methods.medrpm.table[, .N]`

## CircRNAs detected by sample

```{r show_circrnas_2reads_2methods_sample, echo=FALSE, include=TRUE}
circ.2reads.at.least <- circrnas.gtf[RPM > 0 , .N, by = .(sample_id, gene_id)][, .(above_min_reads = .N), by = sample_id]
colnames(circ.2reads.at.least)[colnames(circ.2reads.at.least) == "above_min_reads"] <- paste0("With at least ", min_reads, " reads")

coln <- paste0("With at least ", min_reads, " reads and ", min_methods, " methods")
circ.2reads.at.least.2methods.at.least <- circrnas.morethan2methods.medrpm[, .N, 
                                                                           by = .(sample_id, circ_id)][
                                                                             , .(reliable = .N),
                                                                             by = sample_id]
colnames(circ.2reads.at.least.2methods.at.least)[colnames(circ.2reads.at.least.2methods.at.least) == "reliable"] <- coln
reliable.circ.tab <- merge(circ.2reads.at.least, circ.2reads.at.least.2methods.at.least)
kable(reliable.circ.tab)
```

```{r plot_circrnas_2reads_2methods_sample, echo=FALSE, include=TRUE}
reliable.circ.tab.long <- melt(data = reliable.circ.tab, id.vars = "sample_id", 
                               variable.name = "Filter", value.name = "count")

reliable.circ.barplot <- ggplot(data = reliable.circ.tab.long, 
                                mapping = aes(x = sample_id, y = count,
                                              fill = sample_id)) + 
  geom_bar(stat = "identity", width = .7, colour="black", position = "dodge") + 
  geom_text(stat='identity', aes(label= count), vjust=-0.3, size=default.font.size*(5/14)) +
  facet_wrap(facets = ~ Filter) + ggtitle("CircRNAs per sample") + xlab("") + ylab("Number of circRNAs") + 
  theme_bw()
## compute limits to remove space from x-axis. Add a 10% increase to the top limit
reliable.circ.barplot.limits <- c(0, max(reliable.circ.tab.long$count)*1.1)
reliable.circ.barplot <- reliable.circ.barplot + guides(fill = FALSE) +
  theme(text = element_text(size=default.font.size), axis.text.x = element_text(angle=45, hjust=1)) + 
  scale_y_continuous(expand = c(0, 0), limits = reliable.circ.barplot.limits)
reliable.circ.barplot

# reliable.circ.barplot <- ggplot(data = reliable.circ.tab, 
#                                 mapping = aes(x = sample_id, y = `With at least 2 reads and 2 methods`,
#                                               fill = sample_id)) + 
#   geom_bar(stat = "identity", width = .7, colour="black") + 
#   geom_text(stat='identity', aes(label=`With at least 2 reads and 2 methods`), vjust=-0.3, size=default.font.size*(5/14)) +
#   ggtitle("CircRNAs per sample") + xlab("") + ylab("Reliable circRNAs") + 
#   theme_bw()
# ## compute limits to remove space from x-axis. Add a 10% increase to the top limit
# reliable.circ.barplot.limits <- c(0, max(circ.2reads.at.least.2methods.at.least$`With at least 2 reads and 2 methods`)*1.1)
# reliable.circ.barplot <- reliable.circ.barplot + guides(fill=FALSE) + 
#   theme(text = element_text(size=default.font.size), axis.text.x = element_text(angle=45, hjust=1)) + 
#   scale_y_continuous(expand = c(0, 0), limits = reliable.circ.barplot.limits)
# reliable.circ.barplot

```

## CircRNAs expressed per gene, considering genes expressing circRNAs

The following analysis regards the subset of `r circrnas.morethan2methods.medrpm.table[, .N]` detected with at least `r min_reads` reads and by at least `r min_methods` methods (__reliable circRNAs__). 

```{r circrnas_per_gene, echo=FALSE, include=TRUE}
## Number of circRNAs per gene
setkey(combined_gtf, gene_id)
circRNAs.per.gene <- combined_gtf[gene_id != ".", 
                                  .(circ_id = paste0(chr, ":", circ_start,  "-", circ_end, ":", strand),
                                    gene_name), 
                                  by = .(gene_id)] #, chr, circ_start,  circ_end, strand

#setkey(circRNAs.per.gene, circ_id)
## circRNAs.per.gene[circrnas.morethan2methods.medrpm.table[, .(circ_id)]]
n.circ.per.gene <- merge(circRNAs.per.gene, 
                         circrnas.morethan2methods.medrpm.table[, .(circ_id)], 
                         all = F, by = "circ_id")
n.circ.per.gene <- n.circ.per.gene[, .(n_circRNAs = length(unique(circ_id)), 
                     circ_ids = paste0(unique(circ_id), collapse = "|")), 
                 by = .(gene_id, gene_name)][order(-n_circRNAs)]

## save circRNAs per gene
circRNAs_per_gene.file <- file.path(results.dir, "circRNAs_per_gene.csv")
write.csv(x = n.circ.per.gene, 
          file = circRNAs_per_gene.file, 
          row.names = F)

circRNAs.per.gene <- n.circ.per.gene[, .(gene_id, gene_name, circRNAs_per_gene = n_circRNAs)]
# circRNAs.per.gene <- circRNAs.per.gene[circrnas.morethan2methods.medrpm.table[, .(circ_id)]][, .(circRNAs_per_gene = .N), by = gene_id]
```

Number of genes overlapping the `r circrnas.morethan2methods.medrpm.table[, .N]` reliable circRNAs: `r circRNAs.per.gene[, .N]`

Number of circRNAs expressed by each gene, with the respective circRNAs IDs, are raported in file `r circRNAs_per_gene.file` (only genes expressing circRNAs are reported).

Distribution of number of circRNAs per gene
```{r plot_circrnas_per_gene, echo=FALSE, include=TRUE}
## quartiles
kable(data.frame(Value = unclass(summary(circRNAs.per.gene$circRNAs_per_gene))))

## barplot
circRNAs.per.gene.plot <- ggplot(data = circRNAs.per.gene, mapping = aes(x = factor(circRNAs_per_gene))) + 
  geom_bar(stat = "count") + ggtitle("CircRNAs per gene") + xlab("Number of circRNAs") + ylab("Number of genes") + 
  theme_bw()
## compute limits to remove space from x-axis. Add a 10% increase to the top limit
circRNAs.per.gene.plot.limits <- c(0, max(circRNAs.per.gene[, .N, by=circRNAs_per_gene][, N])*1.1)
circRNAs.per.gene.plot <- circRNAs.per.gene.plot + 
  theme(text = element_text(size=default.font.size), axis.text.x = element_text(angle=90, hjust=1)) + 
  scale_y_continuous(expand = c(0, 0), limits = circRNAs.per.gene.plot.limits)
circRNAs.per.gene.plot

# # compute lower and upper whiskers
# ylim_circRNAs.per.gene = boxplot.stats(circRNAs.per.gene$circRNAs_per_gene)$stats[c(1, 5)]
# # scale y limits based on ylim1
# circRNAs.per.gene.boxplot <- ggplot(data = circRNAs.per.gene, mapping = aes(x = "", y = circRNAs_per_gene)) + 
#   geom_boxplot(notch = T, varwidth = T, width = 0.3) + coord_cartesian(ylim = ylim_circRNAs.per.gene*1.05) + xlab("") + 
#   ggtitle("CircRNAs per gene (outliers removed)") + theme_bw() + 
#   theme(text = element_text(size=default.font.size))
# 
# multiplot(circRNAs.per.gene.plot, circRNAs.per.gene.boxplot, cols = 2)
```


## CircRNA categories detected 

Considering reliable circRNAs, the table below indicate the mumber of circRNAs falling in each annotation class according to backsplice start and end positions in relation to annotated exon or introns from overlapping genes.   
NB: circRNA backsplice positions may intersect exon/introns of several genes if multiple genes overlap.

```{r show_circrnas_categories, echo=FALSE, include=TRUE}
circrna.categories.tab <- circrna.rpms.genes[, .N, by = .(circ_id, label)]
setkey(circrna.categories.tab, circ_id)
circrna.categories.tab <- circrna.categories.tab[circrnas.morethan2methods.medrpm.table[, .(circ_id)]][, .N, by = label][order(-N), .(Category = label, "Tot. circRNAs" = N)]
kable(circrna.categories.tab, caption = "")
```

# CircRNA and gene expression

CircRNA expression by sample have been saved in file `r file.path(results.dir, "circRNA_expression_per_sample.csv")`

Gene expression is reported in file `r gene_fpkm_file`.

## Expressed genes

```{r show_gene_expressed_by_sample}
# get gene expressions
if(file.info(gene_fpkm_file)$size > 0){
    if(gene_meta_file != ""){
        gene_meta <- fread(gene_meta_file)
        
        gene_fpkm <- as.data.table(read.table(file = gene_fpkm_file, header = T, sep = "\t", dec = ".", 
                                              colClasses = c("character", "factor", "numeric", 
                                                             "numeric", "numeric", "numeric", 
                                                             "numeric", "character", "character")))
        expressed_genes <- gene_fpkm[, FPKM := round(FPKM, digits = 8)][FPKM > 0]
        gene_meta[, sample_id := sub("cuffquant/(.*)_hisat2.*", "\\1", file)]
        setkey(gene_meta, condition, replicate_num)
        setkey(expressed_genes, condition, replicate)
        expressed_genes <- expressed_genes[gene_meta[, .(sample_id, condition, replicate_num)]]
    }else{
        gene_fpkm <- fread(gene_fpkm_file)
        expressed_genes <- gene_fpkm[, FPKM := round(FPKM, digits = 8)][FPKM > 0]
        expressed_genes$sample_id <- unique(circrnas.gtf$sample_id)
    }
    
    expressed_genes.by.sample <- expressed_genes[, .N, by = .(sample_id, tracking_id)][, .("Tot. gene expressed" = .N), by = sample_id]
    kable(rbind(expressed_genes.by.sample, list("Non-redundant", expressed_genes[, .N, by = tracking_id][, .N])))
}else{
    "Linear gene expression analysis not enabled"
}
```

```{r plot_gene_expressed_by_sample}
if(file.info(gene_fpkm_file)$size > 0){
expressed_genes.by.sample.plot <- ggplot(data = expressed_genes.by.sample, 
                                         mapping = aes(x = sample_id, y = `Tot. gene expressed`,
                                                       fill = sample_id)) + 
  geom_bar(stat = "identity", width = .7, colour="black") + 
  geom_text(stat='identity', aes(label = `Tot. gene expressed`), vjust=-0.3, size=default.font.size*(5/14)) +
  ggtitle("Gene expressed per sample") + xlab("") + ylab("Number of genes") + 
  theme_bw()
## compute limits to remove space from x-axis. Add a 10% increase to the top limit
expressed_genes.by.sample.plot.limits <- c(0, max(expressed_genes.by.sample$`Tot. gene expressed`)*1.1)
expressed_genes.by.sample.plot <- expressed_genes.by.sample.plot + guides(fill = FALSE) +
  theme(text = element_text(size=default.font.size), axis.text.x = element_text(angle=45, hjust=1)) + 
  scale_y_continuous(expand = c(0, 0), limits = expressed_genes.by.sample.plot.limits)
expressed_genes.by.sample.plot
}
```

## CircRNA expression distribution by sample

```{r show_reliable_circrnas_expression, echo=FALSE, include=TRUE}
kable(sapply(X = split(circrnas.morethan2methods.medrpm, circrnas.morethan2methods.medrpm$sample_id), 
       FUN = function(x){summary(x$medRPM)}), caption = "CircRNA expression (RPM) distribution")
```

```{r plot_reliable_circrnas_expression, echo=FALSE, include=TRUE}
## boxplot expression per sample: circRNAs

# notch:	Notches are used to compare groups; if the notches of two boxes do not overlap, 
# this suggests that the medians are significantly different.
# varwidth: boxes are drawn with widths proportional to the square-roots of the number 
# of observations in the groups (possibly weighted, using the weight aesthetic).

circrna.xpr.per.sample.boxplot <- ggplot(data = circrnas.morethan2methods.medrpm, 
                                         mapping = aes(x = sample_id, y = medRPM, fill = sample_id)) + 
  geom_boxplot(notch = T, varwidth = T) + scale_y_log10(labels = scales::comma) + xlab(NULL) + ylab("RPM") + 
  ggtitle("CircRNA expression\nper sample") + theme_bw()
circrna.xpr.per.sample.boxplot <- circrna.xpr.per.sample.boxplot + guides(fill=FALSE) + 
  theme(text = element_text(size=default.font.size), axis.text.x = element_text(angle=45, hjust=1))

```

## Gene expression distribution by sample

```{r show_gene_expression}
if(file.info(gene_fpkm_file)$size > 0){
kable(sapply(X = split(expressed_genes, expressed_genes$sample_id), 
       FUN = function(x){summary(x$FPKM)}), caption = "Gene expression (FPKM) distribution")
}
```

## CircRNA and gene expression plots

### Expression distribution

Notes on boxplot decorations:  

- notches are used to compare groups; if the notches of two boxes do not overlap, this suggests that the medians are significantly different.  
- boxes are drawn with widths proportional to the square-roots of the number of observations in the groups  

```{r plot_gene_expression}
if(file.info(gene_fpkm_file)$size > 0){
## boxplot expression per sample: genes
gene.xpr.per.sample.boxplot <- ggplot(data = expressed_genes, 
                                      mapping = aes(x = sample_id, y = FPKM, fill = sample_id)) + 
  geom_boxplot(notch = T, varwidth = T) + scale_y_log10(labels = scales::comma) + xlab(NULL) + ylab("FPKM") + 
  ggtitle("Gene expression\nper sample") + theme_bw()
gene.xpr.per.sample.boxplot <- gene.xpr.per.sample.boxplot + guides(fill=FALSE) + 
  theme(text = element_text(size=default.font.size), axis.text.x = element_text(angle=45, hjust=1))
}
```


```{r plot_circ_gene_expr}
if(file.info(gene_fpkm_file)$size > 0){
multiplot(circrna.xpr.per.sample.boxplot, gene.xpr.per.sample.boxplot, cols = 2)
}else{
    circrna.xpr.per.sample.boxplot
}
```

### Cumulative expression

```{r cumulative_expression, include=FALSE}
cum_by_sample <- circrnas.morethan2methods.medrpm[order(sample_id, -medRPM), 
                                                  .(csum = cumsum(medRPM), seqid = seq_along(.I), circ_id), 
                                                  by = .(sample_id)]
cum_by_sample[, cond_tot := max(csum), by = .(sample_id)][, perc_csum := csum/cond_tot]

if(file.info(gene_fpkm_file)$size > 0){
cum_gene_by_sample <- expressed_genes[order(sample_id, -FPKM), 
                                      .(csum = cumsum(FPKM), seqid = seq_along(.I), tracking_id), 
                                      by = .(sample_id)]
cum_gene_by_sample[, cond_tot := max(csum), by = .(sample_id)][, perc_csum := csum/cond_tot]
}
```

```{r cumulative_expression_box, fig.width=12, fig.height=4}
## Cumulative expression plot: circRNAs
circrna.cumsum.per.sample <- ggplot(data = cum_by_sample, mapping = aes(x = seqid, y = perc_csum, color = sample_id)) +
  geom_line() + ggtitle("CircRNAs cumulative expression") + xlab("Number of circRNAs") +
  ylab("Cumulative expression") + scale_y_continuous(labels = scales::percent) + scale_color_discrete("Sample") + theme_bw()
circrna.cumsum.per.sample <-circrna.cumsum.per.sample +
  theme(text = element_text(size=default.font.size), legend.position = "bottom")

if(file.info(gene_fpkm_file)$size > 0){
## Cumulative expression plot: genes
gene.cumsum.per.sample <- ggplot(data = cum_gene_by_sample, mapping = aes(x = seqid, y = perc_csum, color = sample_id)) +
  geom_line() + ggtitle("Gene cumulative expression") + xlab("Number of genes") +
  ylab("Cumulative expression") + scale_y_continuous(labels = scales::percent) +
  scale_color_discrete("Sample") + theme_bw()
gene.cumsum.per.sample <- gene.cumsum.per.sample +
  theme(text = element_text(size=default.font.size), legend.position = "bottom")

multiplot(circrna.cumsum.per.sample, gene.cumsum.per.sample, cols=2)
}else{
    circrna.cumsum.per.sample
}


```

## Correlation of circRNA and gene expression 

```{r correlation, echo=FALSE, include=FALSE}

#median circRNA expression (circrna detected by at least 2 methods)
circ_xpr_by_sample <- circrna.rpms.genes[circrnas.morethan2methods][
    medRPM>0, .(medRPM), by = .(circ_id, sample_id)]

# select only gene and circRNAs that are overlapping
setkey(circ_to_gene, label)
circrna_genes <- unique(circ_to_gene[label == "exonic" | label == "intronic", 
                                     .(circ_id = paste0(chr, ":", circ_start, "-", circ_end, ":", strand), 
                                       gene_id, gene_name)])
genic_circrnas <- merge(circrna_genes, circ_xpr_by_sample, all = F, by = "circ_id")

if(file.info(gene_fpkm_file)$size > 0){    
    
    # avg gene expression
    gene_xpr_by_sample <- expressed_genes[, .(gene_id = tracking_id, sample_id, FPKM)]
    
    genic_circrnas <-merge(genic_circrnas, gene_xpr_by_sample, all.x = T, all.y = F, by = c("gene_id", "sample_id"))
    
    # compute coorelations (matrix?)
    genic_circrnas_corr <- genic_circrnas[, .(corr = cor(medRPM, FPKM, method = "spearman")), 
                                          by = .(circ_id, gene_id)]
    
    circrna_gene_xpr_bysample_corr <- genic_circrnas
    if(nrow(genic_circrnas) > 0){
        # put correlations in tabular format
        circrna_gene_xpr_bysample <- dcast(genic_circrnas, 
                                           formula = circ_id + gene_id + gene_name~ sample_id, 
                                           value.var = c("FPKM", "medRPM"), 
                                           fill = 0)
        circrna_gene_xpr_bysample_corr <- merge(circrna_gene_xpr_bysample, genic_circrnas_corr, by = c("circ_id", "gene_id"))
    }
    #save table
    write.csv(x = circrna_gene_xpr_bysample_corr, 
              file = file.path(results.dir, "circRNA_gene_expression_correlation.csv"), 
              row.names = F)
}else{
    circrna_gene_xpr_bysample_corr <- data.table(mock = numeric())
}
```

Expressed circRNAs | Exonic or intronic circRNAs
------------------: | ---------------------------:
`r circ_xpr_by_sample[, .N, by = circ_id][, .N]` | `r genic_circrnas[, .N, by = circ_id][, .N]`

In total, `r ifelse(nrow(circrna_gene_xpr_bysample_corr) > 0, circrna_gene_xpr_bysample_corr[! is.na(corr)][, .N], 0)` correlations were computed involving `r ifelse(nrow(circrna_gene_xpr_bysample_corr) > 0, circrna_gene_xpr_bysample_corr[! is.na(corr)][, .N, by = circ_id][, .N], 0)` circRNAs and 
`r ifelse(nrow(circrna_gene_xpr_bysample_corr) > 0, circrna_gene_xpr_bysample_corr[! is.na(corr)][, .N, by = gene_id][, .N], 0)` genes.  

Correlation values were saved in file `r file.path(results.dir, "circRNA_gene_expression_correlation.csv")`

```{r corr_density_plot}
if(file.info(gene_fpkm_file)$size > 0){
# plot correlations
corr.density.plot <- ggplot(data = circrna_gene_xpr_bysample_corr, mapping = aes(x = corr)) + geom_density()
corr.density.plot <- corr.density.plot + theme_bw() + ggtitle("CircRNA-gene expression correlation distribution") + 
  xlab("Spearman correlation") + ylab("Density") + 
  theme(text = element_text(size=default.font.size)) + scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) + 
  scale_x_continuous(expand = c(0, 0), limits = c(-1, 1))
corr.density.plot

kable(data.frame("Spearman correlation" = unclass(summary(circrna_gene_xpr_bysample_corr$corr))), 
      caption = "Correlations distribution")

kable(data.frame("Spearman correlation" = unclass(summary(circrna_gene_xpr_bysample_corr[corr >= 0]$corr))), 
      caption = "Positive correlations distribution")

kable(data.frame("Spearman correlation" = unclass(summary(circrna_gene_xpr_bysample_corr[corr < 0]$corr))), 
      caption = "Negative correlations distribution")
}
```

```{r correlations_box, fig.width=12, fig.height=4}
if(file.info(gene_fpkm_file)$size > 0){
    cor.boxplot <- ggplot(data = circrna_gene_xpr_bysample_corr) + 
        geom_boxplot(aes(x = "", y = corr), notch = T, width = .5) + 
        theme_bw() + xlab("") + 
        ylab("Correlation") + ggtitle("Correlation\ndistribution")
    
    cor.abs.boxplot <- ggplot(data = circrna_gene_xpr_bysample_corr[
        , .(pos_corr = ifelse(corr > 0, "Positive", "Negative"), corr)][!is.na(corr)]) + 
        geom_boxplot(aes(x = "", y = abs(corr), fill = pos_corr), notch = T, varwidth = T)  + 
        scale_fill_manual(values = c("blue", "red")) + facet_wrap(facets = ~ pos_corr) +
        theme_bw() + guides(fill = FALSE) + xlab("") + ylab("Abs(correlation)") + 
        ggtitle("Absolute negative and positive\ncorrelation distributions")
    
    multiplot(cor.boxplot, cor.abs.boxplot, cols = 2)
    #make HTML version of corralation table
}
```


# Session info

This page was generated with the following packages version

```{r session, echo=FALSE}
sessionInfo()
```