---
title: 'CirComPara: differential gene expression with Cuffdiff'
date: "`r Sys.Date()`"  
output: 
  html_document: 
    keep_md: no
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: 
      collapsed: false
      smooth_scroll: true
    theme: "readable"
---

```{r global_options, include=FALSE}
## variables to be defined in parent frame
# results.dir <- "analysis/cuffdiff"
# meta.file <- "analysis/cuffdiff/read_groups.info"
# gene.de.file <- "analysis/cuffdiff/gene_exp.diff"

dir.create(path = results.dir, showWarnings = F, recursive = T)
knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 6, 
                      dev = c('svglite', 'png'), 
                      fig.pos = "center",
                      fig.path = file.path(results.dir, paste0('Figs_deg_cuffdiff', .Platform$file.sep)),
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)

knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = results.dir)

library(data.table)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(knitr)
library(VennDiagram) ## to plot Venn diagram
library(DT)
```

```{r load data, include=FALSE}
if(meta.file != ""){
    gene_meta <- fread(meta.file, colClasses = c("character", "factor", "character", 
                                                  "numeric", "numeric", "numeric", 
                                                  "numeric"))
    gene_meta[, sample_id := sub("cuffquant/(.*)_hisat2.*", "\\1", file)]
    
    intgroup.dt <- gene_meta[, .(sample_id, condition)][order(sample_id)]
    intgroup.dt[, color_luminance := 100-(10*1:.N), by = condition]
    intgroup.dt <- merge(intgroup.dt, 
                         intgroup.dt[, .(samples_per_condition = .N), 
                                     by = condition][, hue := brewer.pal(8, 
                                                                         "Set2")[1:length(condition)]][], 
                         by = "condition")[, color := muted(hue, l = color_luminance)][]
}

gene.de <- fread(gene.de.file)
gene.de[, contrast := paste(sample_1, "vs", sample_2)]
```

# Summary

```{r summary samples conditions}
kable(x = gene_meta[, .(sample_id, condition)], row.names = T, caption = "Samples' condition groups")
```

The analysis considered `r gene_meta[, .N, by = sample_id][, .N]` samples grouped in `r gene_meta[, .N, by = condition][, .N]` conditions.  

```{r summary samples per condition}
kable(x = gene_meta[, .(samples = .N), by = condition], row.names = F, caption = "Number of replicates for each condition group")
```

The contrasts considered were `r paste0(unique(gene.de[, contrast]), collapse = ", ")`.  

# Differential expression test counts 

The number of significantly differential expressed genes (DEG) is reported in the table below.  

```{r summary de genes per condition}
kable(x = gene.de[significant == "yes", .(`Tot. DEGs` = .N), by = contrast], 
      row.names = F, 
      caption = "Number of differentially expressed genes for each contrast")
```

The number of DEGs shared among the contrasts is reported below.  

```{r DEG shared}
shared.counts.table <- dcast(gene.de[significant == "yes", ], 
                             formula = test_id ~ contrast, 
                             value.var = "significant")[, -c("test_id"), 
                                                        with = F]
kable(x = crossprod(!is.na(shared.counts.table)), row.names = T, 
      caption = "Number of differentially expressed genes shared among contrasts")
```

```{r DEG venn}
n.contrasts <- length(unique(gene.de[, contrast]))
if(n.contrasts > 1){
    fill.colors <- 1:n.contrasts
    if(n.contrasts < 5){fill.colors <- fill.colors +1}
    if(n.contrasts < 6){
    deg.lists.by.contrast <- split(gene.de[significant == "yes", test_id], 
                                   f = gene.de[significant == "yes", contrast])
    #venn(deg.lists.by.contrast) ## venn using gplots
    vp <- venn.diagram(deg.lists.by.contrast, 
                       fill = fill.colors, 
                       alpha = 0.3, filename = NULL,
                       main = "Differentially expressed genes shared by the contrasts", 
                       cex = 3, main.cex = 2, cat.cex = 1.5)
    grid.draw(vp)
    }else{
        print("Too many contrasts to be plotted by a Venn diagram. Skipped.")
    }
}
```

```{r DEG table}
gene.de.tab <- dcast(gene.de[, 
        .(gene_id = test_id, gene, contrast, `log2(fold_change)`, p_value, 
          q_value, avgFPKM = (value_1 + value_2)/2)], 
      formula = gene_id + gene ~ contrast, 
      value.var = c("log2(fold_change)", "avgFPKM", "q_value", "p_value"))

## save table
gene.de.tab.filename <- file.path(results.dir, "DEG_tests_by_contrast.csv")
write.csv(gene.de.tab, file = gene.de.tab.filename, row.names = F)
```

DEG list by contrast was saved in file `r gene.de.tab.filename`

```{r show DEG table}
## display only DEGs
setkey(gene.de.tab, gene_id)
datatable(data = gene.de.tab[unique(gene.de[significant == "yes", .(test_id)]),], 
          rownames = F, 
              style = "bootstrap", class = "compact display", 
              caption = "Differential tests of gene expression. Only DEGs in at least one contrast are reported", 
              fillContainer = F, autoHideNavigation = T, filter = "top")
```

# Session info
```{r sessionInfo}
sessionInfo()
```