---
title: 'CirComPara: read statistics'
date: "`r Sys.Date()`"  
output: 
  html_document: 
    keep_md: no
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: 
      collapsed: false
      smooth_scroll: true
    theme: "readable"
---

```{r global_options, include=FALSE}
## inherited variables
# read_stats_collect.file 
# linear.mapper 
# circrna.reads.stats.file 
# results.dir
# meta_file

library(data.table)
library(ggplot2)
library(RColorBrewer)
library(scales)

dir.create(path = results.dir, showWarnings = F, recursive = T)
knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 6, 
                      dev = c('svglite', 'png'), 
                      fig.pos = "center",
                      fig.path = file.path(results.dir, paste0('Figs_rs', .Platform$file.sep)),
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load data, echo=FALSE, include=FALSE}

read_stats_collect <- as.data.table(data.frame(line = cbind(scan(file = read_stats_collect.file, 
                                                                 what = "character", 
                                                                 multi.line = T, sep = "\n")), stringsAsFactors = F))
## samples
read_stats_collect$Sample <- gsub(pattern = paste(".*samples", "([^", "]*)", ".*", 
                                                  sep = .Platform$file.sep), 
                                  replacement = "\\1", 
                                  x = read_stats_collect$line)

## preprocessing
read_stats_collect$is.preprocessor <- grepl(pattern = "preprocess", x = read_stats_collect$line)

read_stats_collect[is.preprocessor == TRUE, preprocessor := gsub(pattern = paste("samples/[^", 
                                                                                 "]*/processings/preprocess/([^", 
                                                                                 "]*)",  ".*", 
                                                                                 sep = .Platform$file.sep), 
                                                                 replacement = "\\1", 
                                                                 x = line)]

read_stats_collect[is.preprocessor == TRUE, `:=`(`Raw reads` = as.integer(sub(pattern = ".*Input Read[^:]*: ([0-9]*) .*Surviving.*", 
                                                             "\\1", line)),
                                              `Clean reads` = as.integer(sub(pattern = ".*Input Read[^:]*: [0-9]* [^0-9]* ([0-9]*) .*", 
                                                              "\\1", line)),
                                              Dropped = as.integer(sub(pattern = ".*Dropped: ([0-9]*) .*", 
                                                                  "\\1", line)))]
## linearly unmapped
read_stats_collect[, is.mapper := grepl(pattern = linear.mapper, x = line)]
read_stats_collect[is.mapper == TRUE, `:=`(`Linearly unmapped` = ifelse(grepl(pattern = "aligned[^0]* 0 times", x = line), 
                                                          as.integer(sub(pattern = ".* ([0-9]*) .*aligned[^0]* 0 times", 
                                                              "\\1", x = line)), as.integer(NA)))]

## merge table
read.processing.table <- merge(read_stats_collect[is.preprocessor == TRUE, .(Sample, `Raw reads`, `Clean reads`, Dropped)],
                               read_stats_collect[is.mapper == TRUE, ][!is.na(`Linearly unmapped`), 
                                                                       .(Sample, `Linearly unmapped`)])
read.processing.table[, `Linearly mapped` := `Clean reads`-`Linearly unmapped`]
```

# Read (pre-)processing

| Label             | Description                                                                                            |
|-------------------|--------------------------------------------------------------------------------------------------------|
| Raw reads         | Number of raw reads                                                                                    |
| Clean reads       | Number of reads that passed the quality filtering of pre-processing                                    |
| Linearly mapped   | Number of reads linearly aligned (concordantly, if paired-end) 1 or more times to the reference genome |
| Linearly unmapped | Number of reads NOT linearly aligned to the reference genome: the input reads for backsplice search    |
| Dropped           | Number of reads that did NOT pass the quality filtering of pre-processing                              |

Read counts on read pre-processing and alignment.  

```{r read processing table, echo=FALSE}
knitr::kable(read.processing.table[, .(Sample, `Raw reads`, `Clean reads`, Dropped, `Linearly mapped`, `Linearly unmapped`)], 
             caption = "Per sample read count in each processing step")
```

## Per sample comparison of processing step read counts

```{r Per sample comparison of processing step read counts, echo=FALSE}
if(meta_file != ""){
    meta <- fread(meta_file)
    
    intgroup.dt <- meta[, .(sample, condition)][order(sample)]
    intgroup.dt[, color_luminance := 100-(10*1:.N), by = condition]
    intgroup.dt <- merge(intgroup.dt, 
                         intgroup.dt[, .(samples_per_condition = .N), 
                                     by = condition][, hue := brewer.pal(8, 
                                                                         "Set2")[1:length(condition)]][], 
                         by = "condition")[, color := muted(hue, l = color_luminance)][]
}

read.processing.table.m <- melt(read.processing.table, id.vars = "Sample", 
                                variable.name = "Processing step reads", 
                                value.name = "Read count")
read.processing.table.m$`Processing step reads` <- factor(x = read.processing.table.m$`Processing step reads`, 
                                                          levels = c("Raw reads", "Clean reads", "Linearly mapped", 
                                                                     "Linearly unmapped", "Dropped"), 
                                                          ordered = T)

# ggplot(read.processing.table.m, aes(x=Sample, fill=`Processing step reads`, y=`Read count`)) +
#     geom_bar(stat = "identity", position = "dodge")
ggplot(read.processing.table.m, aes(x=`Processing step reads`, y=`Read count`)) + 
    geom_bar(stat = "identity", aes(fill = `Processing step reads`)) + 
    facet_wrap(facets = ~ Sample, scales = "free_y") + 
    theme(legend.position="none", axis.text.x = element_text(angle=45, hjust=1))
```

## Comparison of processing step read counts
```{r Comparison of processing step read counts, echo=FALSE}

read.processing.boxplot <- ggplot(read.processing.table.m, aes(x=`Processing step reads`, y=`Read count`)) + 
    geom_line(aes(group = Sample, color = Sample)) + 
    geom_boxplot(alpha = .8) + geom_point(aes(color = Sample))
if(meta_file != ""){
    read.processing.boxplot + 
        scale_color_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample]))
}else{
    read.processing.boxplot
}
```

## Comparison of samples, for each processing step
```{r Comparison of samples, for each processing step, echo=FALSE}

read.processing.bars <- ggplot(read.processing.table.m, aes(x=Sample, y=`Read count`, fill = Sample)) +
    geom_bar(stat = "identity") + 
    facet_wrap(facets = ~ `Processing step reads`, scales = "free_y") +
    theme(axis.text.x = element_text(angle=45, hjust=1))
if(meta_file != ""){
    read.processing.bars + 
        scale_fill_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample]))
}else{
    read.processing.bars
}

```

# Circular RNA methods' alignments

```{r Circular RNA methods alignments, echo=FALSE, include=FALSE}
circrna.mappings <- as.data.table(data.frame(scan(file = circrna.reads.stats.file, 
                              what = list("", ""), sep = "\n", blank.lines.skip = T), 
                              stringsAsFactors = F))
colnames(circrna.mappings) <- c("file", "Mapped reads")

circrna.mappings[, `:=`(Sample = sub(pattern = paste(".*samples", "([^", "]*)", ".*", 
                                                     sep = .Platform$file.sep), 
                                     replacement = "\\1", x = file),
                        `CircRNA method` = sub("_out", "", sub(pattern = paste(".*circRNAs", "([^", "]*)", ".*", 
                                                     sep = .Platform$file.sep), 
                                     replacement = "\\1", x = file)),
                        `Mapped reads` = as.integer(`Mapped reads`))]

circrna.mappings.table.m <- rbind(circrna.mappings[, .(Sample, `CircRNA method`, `Mapped reads`)], 
                                  read.processing.table.m[`Processing step reads` == "Linearly unmapped", 
                                                          .(Sample, `CircRNA method` = `Processing step reads`, 
                                                            `Mapped reads` = `Read count`)], use.names = T, fill = T)
```

```{r circrna aligned reads table, echo=FALSE}
circrna.mappings.table <- dcast(data = circrna.mappings.table.m[, .(`Mapped reads`, Sample, `CircRNA method`)], 
                                formula = Sample ~ `CircRNA method`, value.var = "Mapped reads")
knitr::kable(x = circrna.mappings.table, caption = "")
```

## Read aligned by each circRNA method, per sample

```{r plot circrna methods, echo=FALSE}
ggplot(data = circrna.mappings.table.m, aes(x = `CircRNA method`, y = `Mapped reads`)) + 
    geom_bar(stat = "identity") + facet_wrap(facets = ~ Sample)  +
    theme(axis.text.x = element_text(angle=45, hjust=1))
```

## CircRNA method aligned read comparison
```{r plot circrna methods compared, echo=FALSE}
circrna.mappings.boxplot <- ggplot(data = circrna.mappings.table.m,#[`CircRNA method` != "Linearly unmapped"], 
       aes(x = `CircRNA method`, y = `Mapped reads`)) + 
    geom_line(aes(group = Sample, color = Sample)) + 
    geom_boxplot(alpha = .8) + geom_point(aes(color = Sample)) +
    theme(axis.text.x = element_text(angle=45, hjust=1))
if(meta_file != ""){
    circrna.mappings.boxplot + 
        scale_color_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample]))
}else{
    circrna.mappings.boxplot
}
```

```{r save read counts, echo=FALSE}
write.csv(x = merge(read.processing.table, 
      dcast(circrna.mappings[, .(Sample, `CircRNA method`, `Mapped reads`)], 
            formula = Sample ~ `CircRNA method`, value.var = "Mapped reads")),
      file = file.path(results.dir, "processing_and_mapped_read_counts.csv"), row.names = F)
```

# Session info

This page was generated with the following packages version

```{r session, echo=FALSE}
sessionInfo()
```