---
title: 'CirComPara: read statistics'
date: "`r Sys.Date()`"  
output: 
  html_document: 
    keep_md: no
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: 
      collapsed: false
      smooth_scroll: true
    theme: "readable"
editor_options: 
  chunk_output_type: console
---

```{r global_options, include=FALSE}
## inherited variables
# read_stats_collect.file 
# linear.mapper 
# circrna.reads.stats.file 
# results.dir
# meta_file

library(data.table)
library(ggplot2)
library(RColorBrewer)
library(scales)
library(DT)

dir.create(path = results.dir, showWarnings = F, recursive = T)
knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 6, 
                      dev = c('svglite', 'png'), 
                      fig.pos = "center",
                      fig.path = file.path(results.dir, paste0('Figs_rs', .Platform$file.sep)),
                      echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load data, echo=FALSE, include=FALSE}

read_stats_collect <- as.data.table(data.frame(line = cbind(scan(file = read_stats_collect.file, 
                                                                 what = "character", 
                                                                 multi.line = T, sep = "\n")), stringsAsFactors = F))
## samples
read_stats_collect$Sample <- gsub(pattern = paste(".*samples", "([^", "]*)", ".*", 
                                                  sep = .Platform$file.sep), 
                                  replacement = "\\1", 
                                  x = read_stats_collect$line)

## preprocessing
read_stats_collect$is.preprocessor <- grepl(pattern = "preprocess", x = read_stats_collect$line)

read_stats_collect[is.preprocessor == TRUE, preprocessor := gsub(pattern = paste("samples/[^", 
                                                                                 "]*/processings/preprocess/([^", 
                                                                                 "]*)",  ".*", 
                                                                                 sep = .Platform$file.sep), 
                                                                 replacement = "\\1", 
                                                                 x = line)]

read_stats_collect[is.preprocessor == TRUE, `:=`(`Raw reads` = as.integer(sub(pattern = ".*Input Read[^:]*: ([0-9]*) .*Surviving.*", 
                                                             "\\1", line)),
                                              `Clean reads` = as.integer(sub(pattern = ".*Input Read[^:]*: [0-9]* [^0-9]* ([0-9]*) .*", 
                                                              "\\1", line)),
                                              Dropped = as.integer(sub(pattern = ".*Dropped: ([0-9]*) .*", 
                                                                  "\\1", line)))]
## linearly unmapped
read_stats_collect[, is.mapper := grepl(pattern = linear.mapper, x = line)]
read_stats_collect[is.mapper == TRUE, `:=`(`Linearly unmapped` = ifelse(grepl(pattern = "aligned[^0]* 0 times", 
                                                                              x = line), 
                                                                        sub(pattern = ".* ([0-9]*) .*aligned[^0]* 0 times", 
                                                                                       "\\1", x = line), 
                                                                        NA))]
read_stats_collect[, `Linearly unmapped` := as.integer(`Linearly unmapped`)]

## merge table
read.processing.table <- merge(read_stats_collect[is.preprocessor == TRUE, .(Sample, `Raw reads`, `Clean reads`, Dropped)],
                               read_stats_collect[is.mapper == TRUE, ][!is.na(`Linearly unmapped`), 
                                                                       .(Sample, `Linearly unmapped`)])
read.processing.table[, `Linearly mapped` := `Clean reads`-`Linearly unmapped`]

## compute percentages
read.processing.table.perc.1 <- read.processing.table[, lapply(.SD, function(x){100*x/`Raw reads`}),
                                                    by = Sample, .SDcols = c("Clean reads", "Dropped")]
colnames(read.processing.table.perc.1)[2:3] <- sapply(colnames(read.processing.table.perc.1)[2:3], 
                                                      function(x)paste0(x, " (% raw)"))
read.processing.table.perc.2 <- read.processing.table[, lapply(.SD, function(x){100*x/`Clean reads`}),
                                                    by = Sample, .SDcols = c("Linearly unmapped", "Linearly mapped")]
colnames(read.processing.table.perc.2)[2:3] <- sapply(colnames(read.processing.table.perc.2)[2:3], 
                                                    function(x)paste0(x, " (% clean)"))
read.processing.table.perc <- merge(read.processing.table.perc.1, read.processing.table.perc.2, by = "Sample")
# read.processing.table.perc <- read.processing.table[, lapply(.SD, function(x){paste0(x, " (", round(100*x/`Raw reads`, 1), " %)")}),
#                                                     by = Sample, .SDcols = c("Clean reads", "Dropped",
#                                                                              "Linearly unmapped", "Linearly mapped")]
# colnames(read.processing.table.perc)[2:5] <- sapply(colnames(read.processing.table.perc)[2:5], function(x)paste0(x, " (% tot.)"))
# merge(read.processing.table[, .(Sample, `Raw reads`)], read.processing.table.perc)

```

# Read (pre-)processing

| Label             | Description                                                                                            |
|-------------------|--------------------------------------------------------------------------------------------------------|
| Raw reads         | Number of raw reads                                                                                    |
| Clean reads       | Number of reads that passed the quality filtering of pre-processing                                    |
| Linearly mapped   | Number of reads linearly aligned (concordantly, if paired-end) 1 or more times to the reference genome |
| Linearly unmapped | Number of reads NOT linearly aligned to the reference genome: the input reads for backsplice search    |
| Dropped           | Number of reads that did NOT pass the quality filtering of pre-processing                              |

Read counts on read pre-processing and alignment.  

```{r read processing table, echo=FALSE}
DT::datatable(read.processing.table[, .(Sample, `Raw reads`, `Clean reads`, Dropped, `Linearly mapped`, `Linearly unmapped`)],
              rownames = F, 
              style = "bootstrap", class = "compact display", 
              caption = "Per sample read count in each processing step", 
              fillContainer = F, autoHideNavigation = T)

DT::datatable(read.processing.table.perc[, lapply(.SD, round, 1), by = Sample, .SDcols = c("Clean reads (% raw)",
                                                                                        "Dropped (% raw)", 
                                                                                        "Linearly unmapped (% clean)",
                                                                                        "Linearly mapped (% clean)")], 
              rownames = F, 
              style = "bootstrap", class = "compact display", 
              caption = "Percentage with respect to raw reads in each processing step", 
              fillContainer = F, autoHideNavigation = T)
```

## Per sample comparison of processing steps

```{r Per sample comparison of processing step read counts, echo=FALSE}
if(meta_file != ""){
    meta <- fread(meta_file)
    
    intgroup.dt <- meta[, .(sample_id = sample, condition)][order(sample_id)]
    intgroup.dt[, color_luminance := 100-(10*1:.N), by = condition]
    intgroup.dt <- merge(intgroup.dt, 
                         intgroup.dt[, .(samples_per_condition = .N), 
                                     by = condition][, hue := brewer.pal(8, 
                                                                         "Set2")[1:length(condition)]][], 
                         by = "condition")[, color := muted(hue, l = color_luminance)][]
}

read.processing.table.m <- melt(read.processing.table, id.vars = "Sample", 
                                variable.name = "Processing step", 
                                value.name = "Read count")
read.processing.table.m$`Processing step` <- factor(x = read.processing.table.m$`Processing step`, 
                                                          levels = c("Raw reads", "Dropped", 
                                                                     "Clean reads", "Linearly mapped", 
                                                                     "Linearly unmapped"), 
                                                          ordered = T)

# ggplot(read.processing.table.m, aes(x=Sample, fill=`Processing step`, y=`Read count`)) +
#     geom_bar(stat = "identity", position = "dodge")
ggplot(read.processing.table.m, aes(x=`Processing step`, y=`Read count`)) + 
    geom_bar(stat = "identity") + #, aes(fill = `Processing step`)) + 
    facet_wrap(facets = ~ Sample, scales = "free_y") + 
    theme(legend.position="none", axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Read counts") + 
    theme(plot.title = element_text(hjust = 0.5))
```

The following plot compares the proportions of the reads resulting from the processing steps in each sample.  
"Total raw reads" = Linearly mapped + Linearly unmapped + Dropped  
Black borders show "Clean reads" = Linearly mapped + Linearly unmapped  

```{r Per sample comparison of processing step read fraction, echo=FALSE}
## by percentages
read.processing.table.perc.m <- melt(read.processing.table.perc, id.vars = "Sample",
                                variable.name = "Processing step",
                                value.name = "Raw reads %")
read.processing.table.perc.m$`Processing step` <- factor(x = read.processing.table.perc.m$`Processing step`,
                                                          levels = c("Clean reads (% raw)", "Dropped (% raw)",
                                                                     "Linearly mapped (% clean)", "Linearly unmapped (% clean)"),
                                                          ordered = T)
# ggplot(read.processing.table.perc.m, aes(x=`Processing step`, y=`Raw reads %`)) + 
#     geom_bar(stat = "identity", aes(fill = `Processing step`)) + 
#     facet_wrap(facets = ~ Sample) + # + ifelse(grepl("raw", `Processing step`), "Raw", "Clean"), drop = TRUE, scales = "free_x"
#     theme(legend.position="none", axis.text.x = element_text(angle=45, hjust=1)) + 
#     ggtitle("Fraction of raw/clean reads") + 
#     theme(plot.title = element_text(hjust = 0.5))

ggplot(data = read.processing.table.m[`Processing step` %in% c("Linearly mapped", "Linearly unmapped", "Dropped")], 
       aes(x = Sample, y=`Read count`, fill = `Processing step`)) + 
    geom_bar(stat = "identity", position = "fill") + 
    geom_bar(data = read.processing.table.m[`Processing step` %in% c("Dropped", "Clean reads")], 
             stat = "identity", position = "fill", color = "black", fill = alpha("white", 0)) + 
    scale_color_manual(name = "Clean reads", values = c("black")) +
    scale_y_continuous(labels = percent) + ylab("Reads") + 
    ggtitle("Fraction of raw/clean reads") +
    theme(plot.title = element_text(hjust = 0.5))
```

## Comparison of processing step read counts
```{r Comparison of processing step read counts, echo=FALSE}
## by read counts
read.processing.boxplot <- ggplot(read.processing.table.m, aes(x=`Processing step`, y=`Read count`)) + 
    geom_line(aes(group = Sample, color = Sample)) + 
    geom_boxplot(alpha = .8) + geom_point(aes(color = Sample)) + 
    ggtitle("Read counts") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    read.processing.boxplot + 
        scale_color_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    read.processing.boxplot
}

## by percentage
read.processing.perc.boxplot <- ggplot(read.processing.table.perc.m, aes(x=`Processing step`, y=`Raw reads %`)) + 
    geom_line(aes(group = Sample, color = Sample)) + 
    geom_boxplot(alpha = .8) + geom_point(aes(color = Sample)) + 
    ggtitle("Fraction of raw/clean reads") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    read.processing.perc.boxplot + 
        scale_color_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    read.processing.perc.boxplot
}
```

## Comparison of samples, for each processing step

```{r Comparison of samples, for each processing step, echo=FALSE}
read.processing.table.m$Sample <- factor(read.processing.table.m$Sample, 
                                                 levels = unique(meta[order(condition, sample), sample]), 
                                                 ordered = T)
## by read counts
read.processing.bars <- ggplot(read.processing.table.m, aes(x=Sample, y=`Read count`, fill = Sample)) +
    geom_bar(stat = "identity") + 
    facet_wrap(facets = ~ `Processing step`, scales = "free_y") +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Read counts") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    read.processing.bars + 
        scale_fill_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    read.processing.bars
}

## by percentages
read.processing.table.perc.m$Sample <- factor(read.processing.table.perc.m$Sample, 
                                                 levels = unique(meta[order(condition, sample), sample]), 
                                                 ordered = T)
read.processing.perc.bars <- ggplot(read.processing.table.perc.m, 
                                    aes(x=Sample, y=`Raw reads %`, fill = Sample)) +
    geom_bar(stat = "identity") + 
    facet_wrap(facets = ~ `Processing step`, scales = "free_y") +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Fraction of raw/clean reads") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    read.processing.perc.bars + 
        scale_fill_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    read.processing.perc.bars
}

```

# Circular RNA methods' alignments

```{r Circular RNA methods alignments, echo=FALSE, include=FALSE}
circrna.mappings <- as.data.table(data.frame(scan(file = circrna.reads.stats.file, 
                              what = list("", ""), sep = "\n", blank.lines.skip = T), 
                              stringsAsFactors = F))
colnames(circrna.mappings) <- c("file", "Mapped reads")

circrna.mappings[, `:=`(Sample = sub(pattern = paste(".*samples", "([^", "]*)", ".*", 
                                                     sep = .Platform$file.sep), 
                                     replacement = "\\1", x = file),
                        `CircRNA method` = sub("_out", "", sub(pattern = paste(".*circRNAs", "([^", "]*)", ".*", 
                                                     sep = .Platform$file.sep), 
                                     replacement = "\\1", x = file)),
                        `Mapped reads` = as.integer(`Mapped reads`))]

circrna.mappings.table.m <- rbind(circrna.mappings[, .(Sample, `CircRNA method`, `Mapped reads`)], 
                                  read.processing.table.m[`Processing step` == "Linearly unmapped", 
                                                          .(Sample, `CircRNA method` = `Processing step`, 
                                                            `Mapped reads` = `Read count`)], use.names = T, fill = T)

circrna.mappings.table.perc.m <- merge(circrna.mappings[, .(Sample, `CircRNA method`, `Mapped reads`)], 
                                  read.processing.table.m[`Processing step` == "Linearly unmapped", 
                                                          .(Sample, `Linearly unmapped` = `Read count`)], 
                                  by = "Sample")[, .("Linearly unmapped %" = `Mapped reads`*100/`Linearly unmapped`), 
                                                 by = .(Sample, `CircRNA method`)]
```

```{r circrna aligned reads table, echo=FALSE}
circrna.mappings.table <- dcast(data = circrna.mappings.table.m[, .(`Mapped reads`, Sample, `CircRNA method`)], 
                                formula = Sample ~ `CircRNA method`, value.var = "Mapped reads")
DT::datatable(circrna.mappings.table, 
              rownames = F, 
              style = "bootstrap", class = "compact display", 
              caption = "CircRNA methods' alignment counts", 
              fillContainer = F, autoHideNavigation = T)
#knitr::kable(x = circrna.mappings.table, caption = "")

circrna.mappings.perc.table <- dcast(data = circrna.mappings.table.perc.m[, `Linearly unmapped %` := round(`Linearly unmapped %`, 1)], 
                                formula = Sample ~ `CircRNA method`, value.var = "Linearly unmapped %")
DT::datatable(circrna.mappings.perc.table, 
              rownames = F, 
              style = "bootstrap", class = "compact display", 
              caption = "CircRNA methods' alignment fraction of linearly unmapped reads", 
              fillContainer = F, autoHideNavigation = T)
```

## Read aligned by each circRNA method, per sample

```{r plot circrna methods, echo=FALSE}
## by read counts
ggplot(data = circrna.mappings.table.m, aes(x = `CircRNA method`, y = `Mapped reads`)) + 
    geom_bar(stat = "identity") + facet_wrap(facets = ~ Sample)  +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("CircRNA methods' alignment counts") + 
    theme(plot.title = element_text(hjust = 0.5))

## by percentages
ggplot(data = circrna.mappings.table.perc.m, aes(x = `CircRNA method`, y = `Linearly unmapped %`)) + 
    geom_bar(stat = "identity") + facet_wrap(facets = ~ Sample)  +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Fraction of linearly unmapped reads") + 
    theme(plot.title = element_text(hjust = 0.5))
```

## CircRNA method aligned read comparison
```{r plot circrna methods compared, echo=FALSE}
## by read count boxplot
circrna.mappings.boxplot <- ggplot(data = circrna.mappings.table.m,#[`CircRNA method` != "Linearly unmapped"], 
       aes(x = `CircRNA method`, y = `Mapped reads`)) + 
    geom_line(aes(group = Sample, color = Sample)) + 
    geom_boxplot(alpha = .8) + geom_point(aes(color = Sample)) +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("CircRNA methods' alignment counts") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    circrna.mappings.boxplot + 
        scale_color_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    circrna.mappings.boxplot
}

## by read count barplot per method
circrna.mappings.barplot <- ggplot(data = circrna.mappings.table.m[`CircRNA method` != "Linearly unmapped"],
       aes(x = Sample, y = `Mapped reads`, fill = Sample)) + 
    geom_bar(stat = "identity") + facet_wrap(facets = ~ `CircRNA method`, scales = "free_y") +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("CircRNA methods' alignment counts") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    circrna.mappings.barplot + 
        scale_fill_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    circrna.mappings.barplot
}

```

```{r plot circrna methods percentage compared, echo=FALSE}
## by percentage
circrna.mappings.perc.boxplot <- ggplot(data = circrna.mappings.table.perc.m,
       aes(x = `CircRNA method`, y = `Linearly unmapped %`)) + 
    geom_line(aes(group = Sample, color = Sample)) + 
    geom_boxplot(alpha = .8) + geom_point(aes(color = Sample)) +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Fraction of linearly unmapped reads") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    circrna.mappings.perc.boxplot + 
        scale_color_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    circrna.mappings.perc.boxplot
}

## by percentage barplot per method
circrna.mappings.table.perc.m$Sample <- factor(circrna.mappings.table.perc.m$Sample,
                                               levels = unique(meta[order(condition, sample), sample]), 
                                               ordered = T)
circrna.mappings.perc.barplot <- ggplot(data = circrna.mappings.table.perc.m,
       aes(x = Sample, y = `Linearly unmapped %`, fill = Sample)) + 
    geom_bar(stat = "identity") + facet_wrap(facets = ~ `CircRNA method`) +
    theme(axis.text.x = element_text(angle=45, hjust=1)) + 
    ggtitle("Fraction of linearly unmapped reads aligned by each circRNA method\nSamples compared by method") + 
    theme(plot.title = element_text(hjust = 0.5))
if(meta_file != ""){
    circrna.mappings.perc.barplot + 
        scale_fill_manual(values = setNames(intgroup.dt[, color], nm = intgroup.dt[, sample_id]))
}else{
    circrna.mappings.perc.barplot
}

```

```{r save read counts, echo=FALSE}
write.csv(x = merge(read.processing.table, 
      dcast(circrna.mappings[, .(Sample, `CircRNA method`, `Mapped reads`)], 
            formula = Sample ~ `CircRNA method`, value.var = "Mapped reads")),
      file = file.path(results.dir, "processing_and_mapped_read_counts.csv"), row.names = F)
```

# Session info

This page was generated with the following packages version

```{r session, echo=FALSE}
sessionInfo()
```